Garbage Collection Evaluation Infrastructure for the Cog VM

### 2018-06-15 paper.tex
fix detail

We built six different benchmarks that falls into three categories: growing, stable and shrinking heaps. The benchmarks were built on top of the stable Moose image\footnote{http://www.moosetechnology.org/}.

\paragraph{Growing heap.} The growing heap benchmarks increase the heap size as they are performed, mainly creating objects. We built two benchmarks:
\begin{itemize}
	\item \emph{LoadFromMSE}: parses a mse file into a moose model, effectively loading a graph of object. This is the only benchmark not taking a moose model as a parameter.
	\item \emph{ExpandProperties}: computes all interesting properties of the model and store them, effectively loading a graph of object connected in many points to the original graph. 
\end{itemize}

\begin{code}
GCMooseBench >> benchLoadFromMSE: fileName
	mooseModel := MooseModel new importFromMSEStream: (StandardFileStream readOnlyFileNamed: fileName)

GCMooseBench >>benchExpandProperties
	allProperties :=  mooseModel entityStorage collect: [ :anEntity | 
    		anEntity mooseInterestingEntity complexPropertyPragmas collect: [ :aPragma |
        		[ (anEntity mooseInterestingEntity perform: aPragma methodSelector) ] on: Error do: ['error'] ] ].
\end{code}

\paragraph{Stable heap.} The stable heap benchmarks perform random memory accesses in a heap which size remains approximately the same. We built three benchmarks:

\begin{itemize}
	\item \emph{Pyramid}: computes an overview of the pyramid metrics of the software.
	\item \emph{Symbol}: computes all the symbols used in name. \todo{check}
	\item \emph{Deprecation}: computes all the deprecated code.
\end{itemize}

\begin{code}
GCMooseBench >> benchAnalysisPyramid
	mooseModel overviewPyramidMetrics.

GCMooseBench >> benchAnalysisSymbol
	mooseModel allMethods symbolsUsedInName

GCMooseBench >> benchAnalysisDeprecation
	mooseModel allModelClasses select: [ :each |
    		(each isAnnotatedWith: 'Deprecated') and: [ each clientTypes notEmpty ] ].
\end{code}

@@

\begin{code}
GCMooseBench >> benchRelease
	mooseModel := nil.
	allProperties := nil.
	3 timesRepeat: [Smalltalk garbageCollect].
\end{code}

@@

>>>>

We built four different benchmarks that falls into three categories: growing, stable and shrinking heaps. The benchmarks were built on top of the stable Moose image\footnote{http://www.moosetechnology.org/}.

\paragraph{Growing heap.} The growing heap benchmarks increase the heap size as they are performed, mainly creating objects. We built two benchmarks:
\begin{itemize}
	\item \emph{LoadFromMSE}: parses a mse file into a moose model, effectively loading a graph of object. This is the only benchmark not taking a moose model as a parameter.
	\item \emph{ExpandProperties}: computes all interesting properties of the model, discard properties rarely used or consuming too much memory, cache common properties not using too much memory. It effectively loads a graph of object connected in many points to the original graph. 
\end{itemize}

\paragraph{Stable heap.} The stable heap benchmarks perform random memory accesses in a heap which size remains approximately the same. We built one benchmark:

\begin{itemize}
	\item \emph{ExpandPropertiesWithCache}: computes all interesting properties of the model, common properties with low memory footprint are already cached and other properties are discarded once computed.
\end{itemize}

@@

@@

We evaluated both reference algorithms on the benchmarks built.

\subsection{Moose Benchmarks}

\begin{figure*}[thb]
	\centering
    	\begin{subfigure}[b]{.48\textwidth}
	\includegraphics[width=\linewidth]{figures/load} 
	\subcaption{Release\vspace{0.5cm}}
   	\end{subfigure} \hspace{0.03\textwidth} \begin{subfigure}[b]{.48\textwidth}
	\includegraphics[width=\linewidth]{figures/prop} 
	\subcaption{ExpandProperties\vspace{0.5cm}}
   	\end{subfigure}

	\begin{subfigure}[b]{.48\textwidth}
	\includegraphics[width=\linewidth]{figures/propCache} 
	\subcaption{ExpandPropertiesWithCache}
   	\end{subfigure} \hspace{0.03\textwidth}  \begin{subfigure}[b]{.48\textwidth}
	\includegraphics[width=\linewidth]{figures/release} 
	\subcaption{release}
   	\end{subfigure}
\caption{Moose benchmark results piled up.}
\label{MooseRes1}
\end{figure*}

\begin{figure*}[thb]
\begin{tabular}{|c|c|c|c|c|c|c|c|}
   \hline
  & & Total   & Scavenge  & Total Full  & Compaction  & Initial  & Final \\
 Benchmark & &  Exec. time &  time & GC time &  time &  Heap size & Heap size \\
  & &  (ms) &  (ms) & (ms) &  (ms) &  (Mb) &  (Mb) \\
   \hline
   \multirow{2}{*}{Load} & Sweep 	& 159821 $\pm$ 2218 &	73599 $\pm$ 1281 	& 10863 $\pm$ 726 	&1398 $\pm$ 091 	& 193 $\pm$ 0 	& 959 $\pm$ 9.69 \\
    				    & Compact 	& 162305 $\pm$ 2790 &	72150 $\pm$ 0782 	& 14091 $\pm$ 972 	&4661 $\pm$ 289 	& 193 $\pm$ 0 	& 909 $\pm$ 9.69 \\
   \hline
   \multirow{2}{*}{ExpandProperties} 	& Sweep 		& 232940 $\pm$ 01287 &	55497 $\pm$ 0238 	& 12240 $\pm$ 285 	&1178 $\pm$ 6.66 	& 959 $\pm$ 9.69 & 1888 $\pm$ 0 \\
    				    			& Compact 	& 183314 $\pm$ 13090 &	53165 $\pm$ 2343 	& 12496 $\pm$ 197 	&3775 $\pm$ 0108 	& 909 $\pm$ 9.69 & 1938 $\pm$ 0 \\
   \hline
   ExpandProperties 		& Sweep 		& 510013 $\pm$ 05812 &	64234 $\pm$ 0823 	& 23225 $\pm$ 100 	&2011 $\pm$ 10 	& 1888 $\pm$ 0 	& 1888 $\pm$ 0 \\
   WithCache  			& Compact 	& 408169 $\pm$ 20444 &	64492 $\pm$ 1234 	& 0 $\pm$ 0 		&0 $\pm$ 0 		& 1938 $\pm$ 0 		& 1938 $\pm$ 0 \\
   \hline
   \multirow{2}{*}{Release} 	& Sweep 		& 0606 $\pm$ 16 		&	0 $\pm$ 0 	& 0605 $\pm$ 16 	&316 $\pm$ 12 		& 1888 $\pm$ 0 	& 534 $\pm$ 58 \\
    				    		& Compact 	& 1092 $\pm$ 02 	&	0 $\pm$ 0 	& 1092 $\pm$ 01 		&806 $\pm$ 03 			& 0193 $\pm$ 0 		& 193 $\pm$ 00 \\
   \hline
\end{tabular} 
\caption{Moose benchmark results with standard errors.}
\label{MooseRes2}
\end{figure*}

We evaluated the Moose benchmarks using the reference Mark-Compact and Mark-Sweep algorithm. The results are shown in Figure \ref{MooseRes1} and \ref{MooseRes2}. In the results, the compaction time includes the time spent in the compact phase in the Mark-Compact and in the sweep phase in the Mark-Sweep. Total full GC time is the total time spent in full GCs, including the compaction time. Scavenge time is the time spent in scavenges, \emph{i.e.,} garbage collection of young objects only. Total execution time is the total execution time, including scavenge and full GC time.

\paragraph{Load.} The Load benchmark grows the memory by 750 Mb with Sweep and by 700 Mb with Compact. Compaction time is smaller for the sweep algorithm, making the benchmark a little bit faster in the Mark-Sweep, which is to be expected since Sweep does not need to waste time updating pointers because objects are not moved in memory.

\paragraph{ExpandProperties.} The ExpandProperties benchmark grows the memory by around 1 Gb. Compaction time is a little bit better with the sweep algorithm as for the previous benchmark, but execution time is much worse. We believe this is due to worse locality of objects.

\paragraph{ExpandPropertiesWithCache.} The ExpandPropertiesWithCache benchmark does not increase the overall memory size. Since the compacting GC has compacted the heap before starting the benchmark, there is enough room to allocate the large objects (large objects are allocated directly in old space) required by the benchmark and there is no need for a full GC during the benchmark. However, with the sweep algorithm, there are multiple small available memory chunks which are not big enough to hold all the requested large objects, so a full GC is required. In addition, execution time is worse, as for the previous benchmark.

\paragraph{Release.} The Release benchmark releases all the graphs. With the compacting GC, memory goes back to its original size. With Sweep, multiple memory segments are left with a few live objects, forbidding the OS to reclaim those segments. Hence, the memory decreases drastically, but not to the original size. Compaction time is much faster with Sweep, but as we have discussed, for less efficiency.

\subsection{Sophie section}
